Determinism, Complexity, and Predictability in Computer Performance

### 2014-06-17 response.tex

\section*{Response to the Second Referee}

We thank the reviewer for taking the time to both read our paper and provide us
with valuable feedback. While we agree with several of the reviewers criticisms,
and have taken steps to address them, we disagree that doing so changes the
nature of the paper to the point of requiring a new submission.

\emph{The problem of time series prediction is considered in the paper of J.
Garland et al. Namely, the authors suggest to quantify predictability in advance
(before using any concrete prediction scheme) based on the complexity estimate
(weighted permutation entropy, WPE). They hypothesize that WPE is related to the
best possible prediction error in a simple way. For several benchmark time
series (reflecting variations of a computer performance during execution of
different tasks), they obtain estimates of WPE and prediction error (the least
one over four techniques), and build a linear regression using seven or eight
resulting data points. They suppose that the obtained linear function can be
used to ``predict'' the best prediction error from the model-free WPE estimate.
Thus, the authors finally assume (as could be seen from the last section) that
WPE and optimal prediction error for all systems are related via the same
universal linear function with fixed coefficients. This is what I can understand
from the entire paper, though the problem is not so definitely described in the
Introduction.}

\alert{We have modified the introduction to make our goals more clearly
defined.}

@@

\emph{I think that the possibility to estimate the best prediction error in
advance from a relatively simple WPE characteristic would be interesting and
useful, if it could hold true for a wide class of systems. Such a question would
deserve a careful study as that started in the presented work. However, I have a
number of critical remarks and strongly doubt that the conjecture of the authors
can be valid for a reasonably wide class of systems (not even speaking of ``all
systems''). Other aspects of the paper are not so important, in my view, despite
some of them are fresh and interesting (e.g. taking a computer as an object of
prediction). My critical remarks are listed below.}

As we stated on page 19 of the manuscript, we believe that due to the broad
spectrum of behaviors exhibited by the computer performance data these results
will hold for time series drawn from other physical systems. We are aware that
this claim may not be true, and believe that to be critical follow up work.
\alert{However, we have included here some preliminary results that support our
claim by analyzing a classical random walk, three time series derived from the
classic Lorenz attractor, and the time series used in the SFI prediction
competition.} Though we are not prepared to fully discuss or publish the results
regarding these systems, we hope that the preliminary findings are enough to
justify our claim for the reviewer.

@@

\emph{1) Introduction is too long, but the main problem is not formulated
definitely enough. I see that the main point is the approximately linear (or at
least one-to-one) relationship between WPE and the best prediction error. This
aim is not claimed explicitly in the Introduction. All other aspects of the
manuscript do not seem important enough to be published in a separate article.}

\alert{We have modified the manuscript to make it more clear that our goal is to
derive a criterion for best-case prediction error in a model-free way. We have
also shortened it.}

@@

\emph{2) In particular, two primary findings mentioned on page 2 (Introduction)
are formulated quite vaguely. The first part of the first finding reads ``(i)
complexity of a noisy real-value time series is quantifiable by permutation
entropy''. However, it seems obvious from the previous research, including that
cited by the authors. The authors do not suggest PE or WPE here, they exploit
them as previously known approaches. It was previously known that PE is an
estimate of KS entropy as the authors also state. It was well known that KS
entropy relates to predictability and, in this sense, it is a measure of
complexity (see e.g. the work [G. Boffetta, M. Cencini, M. Falcioni, and A.
Vulpiani, ``Predictability: a way to characterize complexity'' // Physics
Reports 356 (2002) 367–474] which is not cited by the authors). Thus, the first
part of the first finding does not provide any new information. The second part
of the first finding reads ``(ii) complexity of the noisy real-valued time
series is correlated with prediction accuracy of an appropriate predictor''.
However, it seems also obvious and directly follows from the definition of KS-
entropy and its discussion in many papers including [Boffetta et al.] cited
above.}

While we agree that on the surface these claims are both trivial and reasonably
well known in theory and for toy models, our specific intent is that these also
hold and can be practically applied via the WPE in real-world systems with
potentially both dynamical and observational noise. We also greatly appreciated
being informed about Boffetta et al's paper, \alert{and have included a citation
to it in our paper.}

@@

\emph{3) By the way, the authors call ``fully complex'' the white noise process,
i.e. they equate predictability and complexity. However, they do not even
mention another approach to the notion of complexity which ascribes low
complexity to the white noise [the works e.g. by C.R. Shalizi and J. Crutchfield
carried out in the Santa Fe Institute from where one of the authors is]. It
would be appropriate to mention that ambiguity of the notion of complexity if
the latter plays an important role in many formulations of the authors.}

\alert{We have added a short paragraph discussing the differences in notions of
complexity, and clarified our working definition of it as used for this paper.}

### 2014-06-17 concl.tex

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of a time
series.  To operationalize that definition, we use an approximation of
the Kolmogorov-Sinai entropy~\cite{lind95}, estimated using a weighted
version of the permutation entropy of~\cite{bandt2002per}.  This WPE
technique---an ordinal calculation of forward information transfer in
a time series---is ideal for our purposes because it works with
real-valued data and is known to converge to the true entropy value.
Using a variety of forecast models and more than a hundred time-series
data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation
entropy: the higher the WPE, in general, the higher the prediction
error.  There are some exceptions: \svdone is easier to predict than
its WPE value suggests using any of the three forecast strategies
studied here.  Conversely, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.  On the whole, though, signals with
higher weighted permutation entropy are harder to predict.

>>>>

In this paper, we have argued that redundancy~\cite{crutchfield2003} is a useful
definition of the inherent predictability of a time series.  To operationalize
that definition, we use an approximation of the Kolmogorov-Sinai
entropy~\cite{lind95}, estimated using a weighted version of the permutation
entropy of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation of
forward information transfer in a time series---is ideal for our purposes
because it works with real-valued data and is known to converge to the true
entropy value. Using a variety of forecast models and more than a hundred time-
series data sets from a computer-performance experiment, we have shown that
prediction accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  Supporting our
hypothesis, forecast errors for the \col and \svdfive signals are higher than
the corresponding WPEs suggest---except for the nonlinear LMA predictor.  There
are some exceptions: \svdone is easier to predict than its WPE value suggests
using any of the three forecast strategies studied here.  On the whole, though,
signals with higher weighted permutation entropy are harder to predict.

### 2014-06-17 model.tex

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
% On this type of plot, a perfect prediction lies exactly on the
% diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

>>>>

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
    \caption{\svdfive \\ LMA}
    \label{fig:svd5LMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
% On this type of plot, a perfect prediction lies exactly on the
% diagonal.
%
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*}

\begin{table*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{lccc}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
    \begin{sideways}\large random walk\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
    \begin{sideways}\large na\"ive\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
    \begin{sideways}\large ARIMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
    \begin{sideways}\large LMA\end{sideways} &
    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
  \label{fig:forecast-example}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
     \gcc, and \svdfive and all four prediction strategies.
  }
  \label{fig:forecast-example}
\end{table*}

In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of Figures~\ref{fig:colRW} and~\ref{fig:colARIMA} are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For ARIMA, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the ARIMA $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.
